Legion: Expressing Locality and Independence with Logical Regions

### 2012-04-28 doc/sc2012/legion.tex
intro

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is very poor performance.  Thus,
to achieve high performance it is crucial that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} (e.g., {\em read-only} and {\em read-write}, among
others) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Priveleges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with {\em read-only} privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent
accesses to the region.  For example, if tasks $x$ and $y$ access
region $r$ with {\em exclusive} access, then Legion guarantees that the effects of $x$ and
$y$ on $r$ are guaranteed to be the same as if they had executed in
the sequential order expressed by the program. On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either $x$
appears to run entirely before $y$ or vice versa.

Legion programs are written using logical regions, each of which
has one or more {\em physical instances} when the program executes.  Often
logical regions correspond to a single physical instance, but there are common situations
in which it is desirable to have multiple physical copies of a logical region's
contents in different parts of the machine (e.g., to duplicate and distribute read-only
data) or even to have physical instances of a logical region that temporarily diverge
in their contents (e.g., to keep an old copy of a region to break a write-after-read dependency, or to 
allow independent, local reductions that are later combined).

@@

>>>>

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
to achieve high performance it is crucial that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} (e.g., {\em read-only} and {\em read-write}, among
others) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Priveleges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with read-only privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent
accesses to the region.  For example, if tasks $x$ and $y$ access
region $r$ with {\em exclusive} access, then Legion guarantees that the effects of $x$ and
$y$ on $r$ are the same as if they had executed in
the sequential order expressed by the program. On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either task $x$
appears to run entirely before $y$ or vice versa.

Legion programs are written using logical regions, each of which
has one or more {\em physical instances} when the program executes.  Often
logical regions correspond to a single physical instance, but it is also common and useful
to have multiple physical copies of a logical region's contents in different parts of the machine 
to duplicate and distribute read-only data, to keep an old copy of a region to break a 
write-after-read dependency, or to allow independent, local reductions that are later combined.

@@

\item We describe a {\em deferred execution} model that extracts parallelism
even from highly irregular, unbalanced computations.  Instead of blocking,
an operation $o$ that cannot immediately execute because it is dependent
on some other operation $o'$ waits on an event that occurs when $o'$ completes.
Events can be created for the completion of any operation in Legion, and 
the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
we show that dataflow-like deferred execution gives us a significant performance improvement
over builk synchronous global synchronization primitives such as barriers.
